A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-04 graph_syntax.tex
technical edits and theoretical comments

\section{Requirements for New Graph Syntax}
The ultimate goal of the new graph--based syntax presented here is to be able to 
fully describe the general structure of an MDAO problem indpendently of any solution information, 
while still being able to accomodate the more specific case when a solution 
strategy is applied. In order to achieve that goal 
the graph needs to accomodate a number of constructs of MDAO problems:

\begin{itemize}
    \item Analysis tools and connections between them
    \item Design variables, objectives, and constraints
    \item Local and global properties
    \item Coupling between analyses
    \item Multi-fidelity analyses
\end{itemize}

Beyond those basic constructs, there are also three phases of a design process that 
all need to be representable with the new syntax. Firstly there is the initial problem definition
phase where the specific analysis tools and design goals are identified. At the end of this phase, 
a single formal problem formulation is selected specifying design variables, constraints, objectives, 
analysis tools, etc. Lastly some specifc procedure for solving the problem is selected, for example 
picking an MDAO optimization architecture. Using the proposed graph syntax, each of these phases 
can be represented with a related graph.

\begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem  (FPG)
    \item Problem Solution Graph (SPG)
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all 
analysis codes being considered and all possible connections between them also present. The second graph 
is \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and node types to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPGs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPGs that could be reached 
untill only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
different solution strategies. If you consider the size of a graph to be the sum of all its
edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
In other words the FPG is obtained from the MCG by removing nodes and edges, 
and the PSG is obtained from the FPG by adding nodes and edges.

@@

\section{Syntax Definition}
In the rest of this section we present the necessary graph theory fundamentals to 
construct any of the different graphs discussed above. 
The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} we construct $E$ as a set of ordered pairs instead 
of a set of sets. Each ordered pair represents an edge starting at the node 
indicated by the first entry and directed to the node indicated by the second 
entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
the edges directed out are given by $E(v)$ and the edges directed into $v$ are given 
by $E^{-1}(v)$; $E(E(v))$ is denoted as $E^2(v)$, and likewise for additional levels.

\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
\end{IEEEeqnarray*}

@@

\subsection{Node and Edge Types}

There are four node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[expression:] simple algebreic combinations of variables (such as for 
  constraint of objective functions)
\item[driver:] control logic capable of managing iteration
\end{description}

In addition to the four node types, there are three edge types:

\begin{description}
\item[fixed edge:] Fundamental, non-changeable property of a problem, such as 
  connections between related variable nodes. These 
edges can never be removed from a graph. 
\item[explicit edge:] Explicit exchange of information between two nodes. No node 
  can have more than a single incomming explicit edge. These edges are free and
  can be added or removed from the graph. 
\item [implicit edge:] Implicit passing of information from a driver node to a 
  variable node. A single variable node can have many incomming implicit edges. These edges are 
  free and can be added or removed from the graph. 
\end{description}

@@

\subsection{Rules for Nodes and Edges}

The driver node and the implicit edge are only allowed to be pressent in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

\subsection{In-degree and Out-degree}

The \emph{in-degree} of a node is the number of fixed or explicit edges directed in and 
is denoted as $\txt{deg}^-(v)$, and the \emph{out-degree} 
is the number of fixed or explicit edges directed out and it is denoted as $\txt{deg}^+(v)$.
We also define the \emph{upper in-degree limit}

\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower in-degree limit}
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}
These user-specified limits govern the number of explicit edges that may be directed into a 
node for a valid graph. Consider a variable node $v$ with
$\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, that node must have exactly one
incomming explicit edge or the graph is invalid. Two possiblities exist for voilating 
the limit conditions: 
\begin{description}
  \item[hole: ] The number of incomming edges is less than $\txt{deg}_l^-(v)$. 
    \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation}

  \item[collision: ] The number of incomming edges is greater than $ \txt{deg}_u^-(v)$. 
    \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \end{equation}
\end{description}

\section{Graph Representation of MDAO Constructs}

\subsection{Analysis Blocks and Connections}

Analysis tools take in a number input variables, then perform some work to calculate 
the values for their respective outputs. In an MDAO graph, this process is 
represented by an group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig. \ref{f:analysis block}, 
the analysis block contains two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to proper output values. If constructing a weighted graph, the weight calculation edge 
provides the opportunity to computational cost. However, it is also allowed to skip 
this edge and have both incomming and outgoing edges to variables through a single model 
node for a given analysis block. Although analysis blocks are comprised of multiple nodes and edges, since all 
the edges within them are fixed, they become fixed sub-graphs within a graph.

@@

Via the analysis block structure, we can also distinguish between input and ouptut
variable nodes. Inputs are any variable nodes that have an outgoing edge into a model 
node. Conversely, outputs are any variable nodes that have an incoming edge from a model node. 
Note that by this definition, inputs and outputs can only exist when variable nodes poses an 
edge joining them to a model node. In Fig. \ref{f:analysis block} these are labeled out as 
local inputs and local outputs. The variables are local because they are explicitely tied
to the analysis block and have no relevance without it.

@@

\subsection{Design Variables, Objectives, and Constraints}

Design variables, objectives, and constraints all require external decisions 
to be made about a given system to identify them. Consider 
a notional problem given by the graph in Fig. \ref{f:designvars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the output, $x$, of $A$
is connected to the input, $x$, of $B$ x can't actually be a design variable. The inputs $w$ and $y$
have no incomming edges and could each be selected as design variables.

\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

More formally, any input variable that can be defined as a hole by Eqn. \ref{e:hole}
would be considered a design variable in the graph. When working with a MCG or FPG, only fixed and explicit edges are allowed, 
in which case a design variable is simply a variable node with no incomming edges. For a valid PSG 
any design variable node would have at least one (but possible multiple) incomming 
implicit edge from a driver node.

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:designvars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns a simple, explicit relationship between them.

@@

@@

\subsection{Local vs Global Nodes}

A simple definition of a global node is any node that involves information 
from more than one analysis block. Likewise, a local node is any node 
that involves information from only a single analysis block. As stated before, variable 
nodes with fixed edges directed to model nodes (as part of an analysis block) are inherently local 
and similarly model nodes are also local. Expression and driver nodes
are not inherently local or global though. For these, their locality is determined 
by the incomming and outgoing edges they posses. Consider the graph in Fig [insert ref], 
where the expression node $g$ represents some kind of constraint for the optimziation problem. 
Since $g$ only had edges comming from analysis block $B$, it would be considered a local constraint 
on block $B$.

@@

While this simple definition is valid, it glosses over a more subtle aspect of 
some MDAO problems. Consider the graph in Fig. [insert ref], with three analysis 
blocks A, B, and C. The expression node, $g$, is a constraint that requires information 
from $A$ and $B$, so it would be considered global by the previous definition. Since $g$
has incomming edges from block $C$, it would not be global with respect to $C$. 
You could say that $g$ is local to the group $A,B$. In this sense, locality 
becomes a relative term. A given node can be local to a specific sub-graph of a large
problem. This subtle structure can become significant when determining how to partition 
an MDAO problem for use with certain MDAO architectures. Martins and Marriage 
demonstrate how Collaborative Optimization outperforms MultiDisciplinary Feasible
when they separated a problem into two sub-problems with variables, objectives, and 
constraints local to each one\cite{Marriage2008a}.

>>>>

\section{Requirements for New Graph Syntax}
The ultimate goal of the new graph--based syntax presented here is to be able to 
fully describe the general structure of an MDAO problem indpendently of any solution information, 
while still being able to accomodate the more specific case when a solution 
strategy is applied. In order to achieve that goal 
the graph needs to accomodate a number of constructs of MDAO problems: 
\begin{itemize}
    \item Analysis tools and connections between them
    \item Design variables, objectives, and constraints
    \item Local and global properties
    \item Coupling between analyses
    \item Multi-fidelity analyses
\end{itemize}

Beyond those basic constructs, there are also three phases of a design process that 
all need to be representable with the new syntax. Firstly there is the initial problem definition
phase where the specific analysis tools and design goals are identified. At the end of this phase, 
a single formal problem formulation is selected specifying design variables, constraints, objectives, 
analysis tools, etc. Lastly some specifc procedure for solving the problem is selected, for example 
picking an MDAO optimization architecture. Using the proposed graph syntax, each of these phases 
can be represented with the following three graphs:
\begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem  (FPG)
    \item Problem Solution Graph (SPG)
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all 
analysis codes being considered and all possible connections between them also present. The second graph 
is the \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and node types to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPGs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPGs that could be reached 
until only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
different solution strategies. If you consider the size of a graph to be the sum of all of its
edges and nodes then the hourglass shape in Fig. \ref{f:hourglass} illustrates how
the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
In other words the FPG is obtained from the MCG by removing nodes and edges, 
and the PSG is obtained from the FPG by adding nodes and edges.

@@

\section{Syntax Definition}
In the rest of this section we present the necessary graph theory fundamentals to 
construct any of the different graphs discussed above. 
The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
of a set of sets. Each ordered pair represents an edge starting at the node 
indicated by the first entry and directed to the node indicated by the second 
entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
the edges directed out are given by $E(v)$ and the edges directed into $v$ are given 
by $E^{-1}(v)$; $E(E(v))$ is denoted as $E^2(v)$, and likewise for additional levels. 
If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
\end{IEEEeqnarray*}

@@

\subsection{Node and Edge Types}
We now define different types of nodes and edges with specific properties to compose graphs used to represent an MDAO problem. 
There are four node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[expression:] simple algebraic combinations of variables (such as for 
  constraint or objective functions)
\item[driver:] control logic capable of managing iteration
\end{description}

In addition to the four node types, there are three edge types: 
\begin{description}
\item[fixed edge:] Fundamental, non-changeable property of a problem, such as 
  connections between related variable nodes. These 
edges can never be removed from a graph. 
\item[explicit edge:] Explicit exchange of information between two nodes. No node 
  can have more than a single incomming explicit edge ((?)). These edges are free and
  can be added or removed from the graph. 
\item [implicit edge:] Implicit passing of information from a driver node to a 
  variable node. A single variable node can have many incomming implicit edges. These edges are 
  free and can be added or removed from the graph. 
\end{description}

@@

\subsection{Rules for Nodes and Edges}
There are specific rules for the usage of these nodes and edges.
The driver node and the implicit edge are only allowed to be present in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to these restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

\subsection{In-degree and Out-degree}
The \emph{in-degree} of a node is the number of fixed or explicit edges directed in and 
is denoted as $\txt{deg}^-(v)$, and the \emph{out-degree} 
is the number of fixed or explicit edges directed out and is denoted as $\txt{deg}^+(v)$.
We also define the \emph{upper in-degree limit} as
\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower in-degree limit} as
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}
These user-specified limits govern the number of explicit edges that may be directed into a variable
node for a valid graph. Consider a variable node $v$ with
$\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, $v$ must have exactly one
incomming explicit edge or the graph is invalid. Two possiblities exist for voilating 
these limit conditions: 
\begin{description}
  \item[hole: ] The number of incomming edges is less than the lower indegree limit:
    \begin{equation} \txt{deg}^-(v) < \txt{deg}_l^-(v) \label{e:hole} \end{equation} 
  \item[collision: ] The number of incomming edges is greater than the upper indegree limit:
    \begin{equation} \txt{deg}^-(v) > \txt{deg}_u^-(v) \end{equation}
\end{description} 
These violations will be managed in order to seek a working problem formulation.

\section{Graph Representation of MDAO Constructs}
%As discussed in Sec.((intro)), the digraph must be able to describe the individual aspects of an MDAO problem.
\subsection{Analysis Blocks and Connections}
Analysis tools take in a number input variables and then perform some work to calculate 
the values for their respective outputs. In an MDAO graph, this process is 
represented by a group of nodes and edges called an \emph{analysis block}, 
shown in Fig. \ref{f:analysis block}. Within an analysis block each variable 
node represents a single input or output and is connected 
to a single model node via a fixed edge. Note that in Fig. \ref{f:analysis block}, 
the analysis block contains two model nodes, with a single edge connecting them. 
This edge represents the necessary calculations to map given inputs 
to proper output values. If constructing a weighted graph, the weight calculation edge 
provides the opportunity to computational cost. However, it is also allowed to skip 
this edge and have both incomming and outgoing edges to variables through a single model 
node for a given analysis block. Although analysis blocks are comprised of multiple nodes and edges, since all 
the edges within them are fixed, they become fixed sub-graphs within a graph.

@@

Via the analysis block structure, we can also distinguish between input and ouptut
variable nodes, as shown in Fig.~\ref{f:analysis block}. Inputs are any variable nodes that have an outgoing edge into a model 
node. Conversely, outputs are any variable nodes that have an incoming edge from a model node. 
Note that by this definition, inputs and outputs can only exist when variable nodes poses an 
edge joining them to a model node. These variables are local because they are explicitely tied
to the analysis block and have no relevance without it.

@@

\subsection{Design Variables, Objectives, and Constraints}
Design variables, objectives, and constraints all require external decisions 
to be made about a given system to identify them. Consider 
a notional problem given by the graph in Fig. \ref{f:designvars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the variable $x$ is connected from $A$ to $B$,
 $x$ can't actually be a design variable. On the other hand, the inputs $w$ and $y$
have no incomming edges and could each be selected as design variables.
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

More formally, any input variable that can be defined as a hole by Eqn. \ref{e:hole}
would be considered a design variable in the graph. When working with an MCG or FPG, only fixed and explicit edges are allowed, 
in which case a design variable is simply a variable node with no incomming edges. For a valid PSG 
any design variable node would have at least one incomming 
implicit edge from a driver node.

((DP: I think we are struggling to define the exact boundary between a hole and design variable and when one can become the other.
This stems from the fact that they are identical with respect to the graph structure: they are both variable nodes representing local inputs without any edges directed in.
Therefore, I suggest we let the difference be strictly determined by the user--set indegree limits: given a node in question, it is a hole if the lower indegree limit is greater than zero. 
Then, a few things can happen. A) the hole can become irrelevant if the analysis block is not needed. B) the hole can cause the graph to be invalid and the user can change the lower indegree limit or change the maximal connectivity graph to include a new edge (and presumably other new nodes and edges as well). 
C) the user could have set the \emph{upper} indegree limit to be zero, meaning that it is supposed to be a design variable.
Having written this, I feel that (if you agree) we should probably have it in the paper somewhere)).

Objectives and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig.~\ref{f:designvars} may have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality or an equality, e.g., 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns a simple, explicit relationship between them. 
((DP:I think we can probably just say ``alegbraic expression'' and not have to show any, especially since the variables aren't given meaning))

@@

((DP: I think this definition of objective/constraint is disharmoneous with our other node types. I think it should also have local inputs represented by variables nodes))

@@

\subsection{Local vs Global Nodes}
A simple definition of a global node is any node that involves information 
from more than one analysis block. Likewise, a local node is any node 
that involves information from only a single analysis block. As stated before, variable 
nodes with fixed edges directed to model nodes (as part of an analysis block) are inherently local 
and similarly model nodes are also local. Expression and driver nodes
are not inherently local or global though. For these, their locality is determined 
by the incomming and outgoing edges they posses. Consider the graph in Fig [insert ref], 
where the expression node $g$ represents some kind of constraint for the optimziation problem. 
Since $g$ only had edges comming from analysis block $B$, it would be considered a local constraint 
on block $B$.

@@

While this simple definition is valid, it glosses over a more subtle aspect of 
some MDAO problems. Consider the graph in Fig. [insert ref], with three analysis 
blocks A, B, and C. The expression node, $g$, is a constraint that requires information 
from $A$ and $B$, so it would be considered global by the previous definition. Since $g$
has incomming edges from block $C$, it would not be global with respect to $C$. 
You could say that $g$ is local to the group $A,B$. In this sense, locality 
becomes a relative term. A given node can be local to a specific sub-graph of a large
problem. This subtle structure can become significant when determining how to partition 
an MDAO problem for use with certain MDAO architectures. Martins and Marriage 
demonstrate how Collaborative Optimization outperforms MultiDisciplinary Feasible
when they separated a problem into two sub-problems with variables, objectives, and 
constraints local to each one\cite{Marriage2008a}. 
((DP: it is not clear to me how the difference between local and global should change in the process of obtaining the FPG from the MCG. For example,
in the MCG a node representing a global input may have several edges directed out to multiple analysis blocks, but in the FPG all but one have been removed. 
From my point of view, I was thinking that the difference between global and local wasn't a node's current state, but what it has the potential to be. Another way of saying this is: the definition of global vs local shouldn't depend on explicit edges because they are free to be removed, unless we also want a node to be able to change from being global or local. Lastly, a global input in an FPG that only has one edge directed out could be removed and the node it was directed in to could become a design variable.))

((DP: along these lines, what if a design variable is a separate variable node outside of the analysis block. Then it doesn't need any indegree limits and being global or local can be directly tied to the number of edges directed out?))
